Compositional Quantitative Resource Analysis of C Programs

### 2014-06-24 paper/main.tex
fix logic, add explanations

It is for instance possible to derive the following quantitative
triple for a binary-search that holds for all stack metrics $M$.  A
stack metric assigns cost $0$ to all evaluation steps except function
calls.  Before a call $f(x)$, we consume $M_f > 0$ resources and
after the call we return $M_f$ resources that can be used in
subsequent function calls.  So $M_f$ corresponds to the stack-frame
size of the function $f$.
%
\begin{center}
${ (Z = \log_2(h-l)) + Z{\cdot}M_\code{bsearch} \} }$
\code{bsearch(x,l,h)}
${ \{ Z{\cdot}M_\code{bsearch} \} }$

@@

Due to \code{break} and \code{return} statements of Clight, there are
different possible ways to exit a block of code.  We also have to keep
track of the resource specifications of functions.  To account for
this in the logic, our quantitative Hoare triples have the form
$$
\Delta; B; R \vdash
\htriple
  { Q }
  { S }
  { Q' } \; .
$$
The triple $\htriple{ Q }{ S }{ Q' }$ consists of a statement $S$ and
two assertions $Q,Q' : \Assn$.  It corresponds to triples in classic
Hoare logic and the intuitive meaning is as follows.  If $S$ is
executed with starting state $\state$, the empty continuation
$\Kstop$, and at least $P(\state)$ resources available then the
evaluation does not run out of resources and there are at least
$Q(\state')$ resources left if the evaluation terminates in state
$\state'$.  However, the assertion $B : \Assn$ provides the
postcondition for the case in which the code block $S$ is exited by a
\code{break} statement.  So if the execution is terminated in state
$\state'$ with a \code{break} then $B(\state')$ resources are
available.  Similarly, $R : \Z \to \Assn$ is the postcondition for the
case in which the code block $S$ is exited by a \code{return x}
statement.  The integer argument of $R$ is the return value.
Finally, the function context 
$$\Delta {:} \text{Funs} \to \Z^*
\\
((\Val \times \Mem) {\to} \N \cup \{\infty\}) {\times} ((\Val \times \Mem) {\to} \N {\cup} \{\infty\})$$

@@

\begin{figure*}[t]
\begin{mathpar}
%
\Rule{L:Assert}
{ }
{ \Delta; B; R \vdash
\htriple
  { Q \implies \code{istrue}~\sem{e}_\state + M_a }
  { \code{assert}~e }
  { Q }
}
%
\and
\Rule{L:Skip}
{ }
{ \Delta; B; R \vdash
\htriple
  { M_s + Q }{ \code{skip} }{ Q }
}
%
\and
\Rule{L:Break}
{ }
{ \Delta; B; R \vdash
\htriple
  { M_b + B }{ \code{break} }{ Q }
}
%
\and
\Rule{L:Return}
{ }
{ \Delta; B; R \vdash
\htriple
  { M_r + R\,x }{ \code{return}~x }{ Q }
}
%
\and
\Rule{L:Update}
{ }
{ \Delta; B; R \vdash
\htriple
  { \lambda \state.\,M_u + Q\,\state[x \mapsto \sem{e}_\state] }{ x \gets e }{ Q }
}
%
\and
\Rule{L:Call}
{
  \Delta f = \forall z\,\vec v.(P_f\,z\,\vec v, Q_f\,z) \\
  P \models P_f \, y \, (\state(\vec x)) \land A \\
  \forall v.\, (Q_f \, y \, v \land A \models \lambda \state.\, Q \, \state[r \mapsto v])
}
{ \Delta; B; R \vdash
\htriple
  { M_f + P }
  { r \gets f(\vec x) }
  { Q }
}
%
\and
\Rule{L:Loop}
{ \Delta; Q; R \vdash
\htriple
  { I }{ S }{ I + M_l }
}
{ \Delta; B; R \vdash
\htriple
  { I }{ \code{loop}~S }{ Q }
}
%
\and
\Rule{L:Seq}
{ \Delta; B; R \vdash
\htriple{ P }{ S_1 }{ Q' + M_s }
\\
  \Delta; B; R \vdash
\htriple{ Q' }{ S_2 }{ Q }
}
{ \Delta; B; R \vdash
\htriple{ P }{ S_1;S_2 }{ Q }
}
%
\and
\Rule{L:If}
{ \Delta; B; R \vdash
\htriple
  { P \land \code{istrue}~\sem{e}_\state - M_c^1}
  { S_1 }{ Q }
\\
  \Delta; B; R \vdash
\htriple
  { P \land \code{isfalse}~\sem{e}_\state - M_c^2 }
  { S_2 }{ Q }
}
{ \Delta; B; R \vdash
\htriple
  { P }{ \code{if}(e)~S_1~\code{else}~S_2 }{ Q }
}
%
\and
\Rule{L:Weaken}
{
  P \models P' \\
  \Delta; B'; R' \vdash
  \htriple{ P' }{ S }{ Q' } \\
  Q' \models Q \\
  B' \models B \\
  \forall v.\, (R'\,v \models R\,v)
}
{ \Delta; B; R \vdash
\htriple{ P }{ S }{ Q }
}
%
\and
\Rule{L:Frame}
{ \Delta; B; R \vdash
\htriple{ P }{ S }{ Q } \\
  x \in \mathbb Q_0^+
}
{ \Delta; B + x; R + x \vdash
\htriple{ P + x }{ S }{ Q + x }
}
%
\and
\Rule{L:Extend}
{ \forall f\,P_f\,Q_f.\,
  \Delta'f = \forall z\,\vec v.(P_f\,z\,\vec v, Q_f\,z)
  \rightarrow
  \forall y\,\vec v.\,
  (\Delta \cup \Delta'; \bot; Q_f\,y \vdash
    \htriple{ P_f\,y\,\vec v }{ S_f }{ \bot })
\\
  \Delta \cup \Delta'; B; R \vdash
  \htriple{ P }{ S }{ Q }
}
{ \Delta; B; R \vdash
\htriple{ P }{ S }{ Q }
}
%
\end{mathpar}

>>>>

It is for instance possible to derive the following quantitative
triple for a binary-search that holds for all stack metrics $M$.  A
stack metric assigns cost $0$ to all evaluation steps except function
calls.  Before a call $f(x)$, we consume $M_f > 0$ resources and
after the call we return $M_f$ resources that can be used in
subsequent function calls.  So $M_f$ corresponds to the stack-frame
size of the function $f$.
%
\begin{center}
${ \{ (Z = \log_2(h-l)) + Z{\cdot}M_\code{bsearch} \} }$
\code{bsearch(x,l,h)}
${ \{ Z{\cdot}M_\code{bsearch} \} }$

@@

Due to \code{break} and \code{return} statements of Clight, there are
different possible ways to exit a block of code.  We also have to keep
track of the resource specifications of functions.  To account for
this in the logic, our quantitative Hoare triples have the form
$$
\Delta; B; R \vdash
\htriple
  { Q }
  { S }
  { Q' } \; .
$$
The triple $\htriple{ Q }{ S }{ Q' }$ consists of a statement $S$ and
two assertions $Q,Q' : \Assn$.  It corresponds to triples in classic
Hoare logic and the intuitive meaning is as follows.  If $S$ is
executed with starting state $\state$, the empty continuation
$\Kstop$, and at least $P(\state)$ resources available then the
evaluation does not run out of resources and there are at least
$Q(\state')$ resources left if the evaluation terminates in state
$\state'$.  The assertion $B : \Assn$ provides the
postcondition for the case in which the code block $S$ is exited by a
\code{break} statement.  So if the execution is terminated in state
$\state'$ with a \code{break} then $B(\state')$ resources are
available.  Similarly, $R : \Z \to \Assn$ is the postcondition for the
case in which the code block $S$ is exited by a $\code{return}~x$
statement.  The integer argument of $R$ is the return value.
Finally, the function context of judgements that we write $\Delta$ is
a mapping from function names to specification of the form
$$
  \forall z \, \vec v \, v .(P_f \, z \, \vec v, Q_f \, z \, v).
$$
The assertion $P_f \, z \, \vec v$ is the precondition of the
function $f$ and the assertion $Q_f \, z \, v$ is its postcondition.
They are both parameterized by an arbitary logical variable $z$
(which can be a tuple) that helps relating them.  The precondition
is allowed to depend on $\vec v$, the value of all the arguments at
the function invocation.  Similarly, the postcondition depends on
the return value $v$ of the function.

@@

\begin{figure*}[t]
\begin{mathpar}
%
\Rule{L:Assert}
{ }
{ \Delta; B; R \vdash
\htriple
  { Q \implies \code{istrue}~\sem{e}_\state + M_a }
  { \code{assert}~e }
  { Q }
}
%
\and
\Rule{L:Skip}
{ }
{ \Delta; B; R \vdash
\htriple
  { Q }{ \code{skip} }{ Q }
}
%
\and
\Rule{L:Break}
{ }
{ \Delta; B; R \vdash
\htriple
  { M_b + B }{ \code{break} }{ Q }
}
%
\and
\Rule{L:Return}
{ }
{ \Delta; B; R \vdash
\htriple
  { R\,(\state(x)) }{ \code{return}~x }{ Q }
}
%
\and
\Rule{L:Update}
{ }
{ \Delta; B; R \vdash
\htriple
  { \lambda \state.\,M_u + Q\,\state[x \mapsto \sem{e}_\state] }{ x \gets e }{ Q }
}
%
\and
\Rule{L:Call}
{
  \Delta(f) = \forall z\,\vec v\,v.(P_f\,z\,\vec v, Q_f\,z\,v) \\
  P \models P_f \, y \, (\state(\vec x)) \land A \\
  \forall v.\, (Q_f \, y \, v \land A \models \lambda \state.\, Q \, \state[r \mapsto v])
}
{ \Delta; B; R \vdash
\htriple
  { M_f + P }
  { r \gets f(\vec x) }
  { Q - M_r }
}
%
\and
\Rule{L:Loop}
{ \Delta; Q; R \vdash
\htriple
  { I }{ S }{ I + M_l }
}
{ \Delta; B; R \vdash
\htriple
  { I }{ \code{loop}~S }{ Q }
}
%
\and
\Rule{L:Seq}
{ \Delta; B; R \vdash
\htriple{ P }{ S_1 }{ Q' + M_s }
\\
  \Delta; B; R \vdash
\htriple{ Q' }{ S_2 }{ Q }
}
{ \Delta; B; R \vdash
\htriple{ P }{ S_1;S_2 }{ Q }
}
%
\and
\Rule{L:If}
{ \Delta; B; R \vdash
\htriple
  { P \land \code{istrue}~\sem{e}_\state - M_c^1}
  { S_1 }{ Q }
\\
  \Delta; B; R \vdash
\htriple
  { P \land \code{isfalse}~\sem{e}_\state - M_c^2 }
  { S_2 }{ Q }
}
{ \Delta; B; R \vdash
\htriple
  { P }{ \code{if}(e)~S_1~\code{else}~S_2 }{ Q }
}
%
\and
\Rule{L:Weaken}
{
  P \models P' \\
  \Delta; B'; R' \vdash
  \htriple{ P' }{ S }{ Q' } \\
  Q' \models Q \\
  B' \models B \\
  \forall v.\, (R'\,v \models R\,v)
}
{ \Delta; B; R \vdash
\htriple{ P }{ S }{ Q }
}
%
\and
\Rule{L:Frame}
{ \Delta; B; R \vdash
\htriple{ P }{ S }{ Q } \\
  x \in \mathbb Q_0^+
}
{ \Delta; B + x; R + x \vdash
\htriple{ P + x }{ S }{ Q + x }
}
%
\and
\Rule{L:Extend}
{ \Delta \cup \Delta'; B; R \vdash
  \htriple{ P }{ S }{ Q }
\\
  \forall f\,P_f\,Q_f.\,
  \Delta'(f) = \forall z\,\vec v\,v.(P_f\,z\,\vec v, Q_f\,z\,v)
  \rightarrow
  \forall y\,\vec v.\,
  (\Delta \cup \Delta'; \bot; Q_f\,y \vdash
    \htriple{ P_f\,y\,\vec v }{ S_f }{ \bot })
}
{ \Delta; B; R \vdash
\htriple{ P }{ S }{ Q }
}
%
\end{mathpar}
